---
title: "Week02_Lecture"
author: "Yanan Wu"
date: "2025-01-06"
output: 
  html_document:
    df_print: paged
    toc: TRUE
    toc_float: TRUE
---
## Calculating confidence interval

calculate the confidence interval when sample size is 800, sample proportion is 0.34 for a 95% confidence interval

```{r}
library(interpretCI)
x=propCI(n = 800, p = 0.34, alpha = 0.05)
x
```

```{r}
n = 800
p = 0.34

se = sqrt(p*(1-p)/n)
print(c('stand error:', se))

margin <- qnorm(0.975)*se
print(c('margin of error:', margin))

print(c('confidence interval: ', c(p - margin, p + margin)))
```


## statistics of p-value

A researcher claims that 46% of the students at a university are left-handed. A sample of 1267 students shows that 559 are left-handed. Is the researcher’s claim accurate at the 1% significance level?

### 1. State hypothesis

# H_0: p = 0.46
# H_a: p != 0.46

```{r}
count <- 1267  # Number of left-handed students in sample
x_success <- 559  # Total number of students

library(stats)
result <- prop.test(x = x_success, n = count, p = 0.46, alternative = "two.sided", conf.level = 0.9)

# Display the result
print(result)
```
### calculation of x-squared with continuity correction 

```{r}
p_hat = 559/1267
p_h0 = 0.46

x_squared_corrected <- ((abs(p_hat - p_h0) - (0.5 / count))^2) / (p_h0 * (1 - p_h0) / count)
print(x_squared_corrected)


```

## Sampling distribution

Suppose you want to estimate the proportion of university students who regularly use public transportation, but you don’t have access to the entire student population. In this case, you could take a sample of students from different campuses and use the proportion of students in your sample who use public transportation as your best guess for the unknown proportion in the overall student population.


```{r}
stu_size <- 250000000
public_use <- c(rep("public", 0.88 * stu_size),rep("nopublic", 0.12 * stu_size))
# 2. Sample 1000 entries without replacement.
p_hat_values <- c()

# Repeat the sampling process 1000 times
for (i in 1:1000) {
  # Sample 1000 entries without replacement
  sampled_students <- sample(public_use, size = 1000)
  
  # Compute p-hat for the current sample
  p_hat_values <- c(p_hat_values, sum(sampled_students == "public") / 1000)
}

# Display the first few p-hat values
head(p_hat_values)
mean(p_hat_values)
# Plot the sampling distribution of p-hat
hist(p_hat_values, main = "Simulated sample proportion", xlab = "sample_proportion", col = "lightblue", border = "white")
```




```{r}
library(ggplot2)

# Create a data frame with two normal distributions
data <- data.frame(
  x = seq(-5, 8, length.out = 1000),
  y1 = dnorm(seq(-5, 8, length.out = 1000), mean = 0, sd = 1),
  y2 = dnorm(seq(-5, 8, length.out = 1000), mean = 3, sd = 1)
)

# Plot the two normal distributions using ggplot2
ggplot(data, aes(x = x)) +
  geom_line(aes(y = y1, color = "Mean = 0"), size = 1) +
  geom_line(aes(y = y2, color = "Mean = 3"), size = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "blue", size = 1) +
  geom_vline(xintercept = 3, linetype = "dashed", color = "red", size = 1) +
  labs(title = "",
       x = "x",
       y = "Density") +
  scale_color_manual(values = c("Mean = 0" = "blue", "Mean = 3" = "red")) +
  theme_minimal()
```







