---
title: "Spatial Autocorrelation Analysis in R"
author: "Yanan Wu"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    code_folding: show
---
## Why is Spatial Autocorrelation Important?

Understanding spatial autocorrelation is crucial because:

1. It indicates the presence of spatial dependence in your data
2. It can reveal underlying spatial processes
3. It may violate the independence assumption of standard statistical models
4. It can help identify spatial clusters or hotspots

## Types of Spatial Autocorrelation

- **Positive spatial autocorrelation**: Similar values cluster together (high near high, low near low)
- **Negative spatial autocorrelation**: Dissimilar values cluster together (high near low)
- **No spatial autocorrelation**: Values are randomly distributed in space

# 2. Setting Up the Environment

Let's start by installing and loading the necessary packages:

```{r packages, message=FALSE, warning=FALSE}
# Install necessary packages (if not already installed)
packages <- c("sf", "spdep", "tmap", "tmaptools", "spData", "dplyr", 
              "ggplot2", "classInt", "RColorBrewer", "viridis", "knitr")
install.packages(packages[!packages %in% installed.packages()[,"Package"]], 
                 repos = "https://cran.r-project.org/")

# Load the packages
library(tmap)       # For thematic maps
library(tmaptools)  # Additional mapping tools
library(dplyr)      # For data manipulation
library(ggplot2)    # For plotting
library(classInt)   # For classification
library(RColorBrewer) # For color palettes
library(viridis)    # For color palettes
library(knitr)      # For nice table output
library(sp)

# Set tmap to plot mode
# tmap_mode("plot")
```

# 3. Loading and Exploring Spatial Data

For this tutorial, we'll use the `columbus` dataset from the `spData` package. This dataset contains 49 neighborhoods in Columbus, Ohio with various socioeconomic variables, including crime rates, housing values, and income.

```{r load_data, message=FALSE, warning=FALSE}
# Load required packages
library(sf)
library(spdep)
library(spData)

# Load the Columbus polygon shapefile correctly
# columbus_poly <- st_read(system.file("shapes/columbus.shp", package="spData"))
columbus_poly <- sf::st_read(system.file("shapes/columbus.gpkg", package="spData")[1])
# columbus_poly <- st_read(system.file("columbus.shp", package="spData"))
# View the first few rows
head(columbus_poly)

# Create a simple map to visualize the areas
tm_shape(columbus_poly) +
  tm_fill("CRIME", style = "quantile", palette = "Reds", title = "Crime Rate") +
  tm_borders() +
  tm_layout(title = "Crime Rates in Columbus Neighborhoods",
            legend.outside = TRUE)
```
Let's explore the distribution of these variables:

```{r data_exploration}
# Summary statistics
summary(columbus_poly[, c("CRIME")])

# Histograms of key variables
par(mfrow = c(1, 1))
hist(columbus_poly$CRIME, main = "Crime Rate", xlab = "Crime Rate", col = "lightblue")

par(mfrow = c(1, 1))

# Create a series of choropleth maps for key variables
tm_shape(columbus_poly) +
  tm_fill(c("CRIME"), 
          style = "quantile", 
          palette = list("Reds"),
          title = c("Crime Rate")) +
  tm_borders() +
  tm_facets(ncol = 1) +
  tm_layout(legend.outside = TRUE)
```

Looking at the maps, we can visually assess potential spatial patterns. For instance, we might observe that high crime areas tend to cluster together, or that neighborhoods with higher housing values are spatially proximate. However, to quantify these patterns, we need formal spatial autocorrelation analysis.

# 4. Creating Spatial Weights

Before we can analyze spatial autocorrelation, we need to define neighborhood relationships between areas. This is done using a spatial weights matrix.

## Types of Spatial Weights

There are several ways to define neighborhood relationships:

1. **Contiguity-based**: Areas that share boundaries
   - Queen: Areas that share any point (borders or corners)
   - Rook: Areas that share borders only

2. **Distance-based**: Areas within a certain distance of each other
   - Fixed distance band
   - K-nearest neighbors

3. **General weights**: Custom weights based on other criteria

## Queen Contiguity Weights

Let's create queen contiguity weights, where regions are neighbors if they share even a single boundary point:

```{r queen_weights}
# Create queen contiguity weights
queen_nb <- poly2nb(columbus_poly, queen = TRUE)
summary(queen_nb)
print(queen_nb)
# Extract coordinates
columbus_sp <- as(columbus_poly, "Spatial")
coords <- coordinates(columbus_sp)

# Plot the queen contiguity network

plot(columbus_sp, border = "grey")
plot(queen_nb, coords, add = TRUE, col = "red", lwd = 1)
title("Queen Contiguity Neighbors")

# Convert to weights list object
queen_weights <- nb2listw(queen_nb, style = "W")

library(tmap)

tm_shape(columbus_poly) +
  tm_fill("CRIME", style = "quantile", palette = "Reds", title = "Crime Rate") +
  tm_borders() +
  # Highlight the first polygon
  tm_shape(columbus_poly[3, ]) +
  tm_borders(lwd = 3, col = "blue") +
  tm_layout(title = "",
            legend.outside = TRUE)

```

## Rook Contiguity Weights

Alternatively, we can create rook contiguity weights, where regions are neighbors only if they share a border segment:

```{r rook_weights}
# Create rook contiguity weights
rook_nb <- poly2nb(columbus_poly, queen = FALSE)
summary(rook_nb)

# Plot the rook contiguity network
plot(columbus_sp, border = "grey")
plot(rook_nb, coords, add = TRUE, col = "red", lwd = 1)
title("Rook Contiguity Neighbors")

# Convert to weights list object
rook_weights <- nb2listw(rook_nb, style = "W")
```

## Distance-Based Weights

We can also create distance-based weights, where neighborhoods within a certain distance are considered neighbors:

```{r distance_weights}


# Create distance-based neighbors
# First, determine an appropriate distance threshold
coords_dist <- as.matrix(dist(coords))
knn1_dist <- apply(coords_dist, 1, function(x) sort(x)[2])
max_knn1 <- max(knn1_dist)
max_knn1

# Create distance-based neighbors list with a threshold that ensures all areas have at least one neighbor
dist_nb <- dnearneigh(coords, 0, max_knn1 * 1.1)
summary(dist_nb)

# Plot the distance-based neighbor network
plot(columbus_sp, border = "grey")
plot(dist_nb, coords, add = TRUE, col = "green", lwd = 1)
title("Distance-Based Neighbors")

# Convert to weights list object
dist_weights <- nb2listw(dist_nb, style = "W")
```

## K-Nearest Neighbors Weights

Another approach is to use k-nearest neighbors, where each region is connected to its k closest neighbors:

```{r knn_weights}
# Create k-nearest neighbors (k=4)
knn_nb <- knn2nb(knearneigh(coords, k = 4))
summary(knn_nb)

# Plot the k-nearest neighbors network
plot(columbus_sp, border = "grey")
plot(knn_nb, coords, add = TRUE, col = "purple", lwd = 1)
title("K-Nearest Neighbors (k=4)")

# Convert to weights list object
knn_weights <- nb2listw(knn_nb, style = "W")
```

# 5. Global Spatial Autocorrelation

Global spatial autocorrelation measures the overall degree of similarity among spatially close observations. Moran's I is the most common statistic for this purpose.

## Moran's I

Moran's I ranges from -1 to 1:
- Values close to 1 indicate positive spatial autocorrelation (clustering)
- Values close to -1 indicate negative spatial autocorrelation (dispersion)
- Values close to 0 indicate random spatial pattern

Let's calculate Moran's I for crime rates:

```{r morans_i}
# Calculate Moran's I for crime rates using queen weights
moran_crime_queen <- moran.test(columbus_sp$CRIME, queen_weights)
print(moran_crime_queen)
```

## Interpreting Moran's I

The output of the Moran's I test includes:

1. **Moran I statistic**: The actual I value, ranging from -1 to 1
2. **Expected value**: The expected value of I under the null hypothesis of no spatial autocorrelation (usually close to 0)
3. **Variance**: The variance of I
4. **Z-score**: The standard score, indicating how many standard deviations the I statistic is from the expected value
5. **p-value**: The probability of observing the calculated I value (or a more extreme one) under the null hypothesis

A significant positive I value (with a low p-value) indicates positive spatial autocorrelation - similar values tend to cluster together geographically. A significant negative I value indicates negative spatial autocorrelation - dissimilar values tend to cluster together.

## Moran Scatterplot

A Moran scatterplot visualizes the relationship between a variable and its spatial lag (the weighted average of neighboring values). The x-axis represents the standardized variable values, and the y-axis represents the standardized spatial lag values.

```{r moran_scatterplot}
# Create Moran scatterplot for crime rates
moran.plot(columbus_sp$CRIME, queen_weights, 
           labels = as.character(columbus_sp$POLYID), 
           xlab = "Crime Rate (standardized)", 
           ylab = "Spatially Lagged Crime Rate")
```

The Moran scatterplot has four quadrants:
- **High-High (upper right)**: High values surrounded by high values
- **Low-Low (lower left)**: Low values surrounded by low values
- **High-Low (lower right)**: High values surrounded by low values
- **Low-High (upper left)**: Low values surrounded by high values

The slope of the regression line through these points is equivalent to Moran's I.

# 6. Local Spatial Autocorrelation

While global measures tell us about the overall pattern, local measures identify specific clusters or outliers at the local level.

## Local Indicators of Spatial Association (LISA)

LISA statistics, introduced by Luc Anselin, decompose global spatial autocorrelation into contributions from individual locations. They help identify:
- Local clusters (hotspots and coldspots)
- Spatial outliers

Let's calculate local Moran's I:

```{r lisa}
# Calculate local Moran's I for crime rates
lisa_crime <- localmoran(columbus_sp$CRIME, queen_weights)
head(lisa_crime)

# Add local Moran statistics to our spatial data
columbus_sp$local_moran <- lisa_crime[, 1]  # Local Moran's I statistic
columbus_sp$local_moran_p <- lisa_crime[, 5]  # P-value

columbus_sf <- st_as_sf(columbus_sp)

# Map the local Moran's I values
tm_shape(columbus_sf) +
  tm_fill("local_moran", style = "jenks", palette = "-RdBu",
          midpoint = 0, title = "Local Moran's I") +
  tm_borders() +
  tm_layout(title = "Local Spatial Autocorrelation of Crime Rates",
            legend.outside = TRUE)

# Map the significance of local Moran's I
tm_shape(columbus_sf) +
  tm_fill("local_moran_p", style = "fixed", 
          breaks = c(0, 0.01, 0.05, 0.1, 1),
          palette = "-Greens", title = "p-value") +
  tm_borders() +
  tm_layout(title = "Significance of Local Moran's I",
            legend.outside = TRUE)
```

## Creating a LISA Cluster Map

LISA cluster maps classify locations into significant clusters and outliers:

```{r lisa_cluster}
# Create a function to classify LISA clusters
lisa_clusters <- function(var, lisa, p_value, p_threshold = 0.05) {
  # Standardize the variable
  z_var <- scale(var)
  
  # Calculate spatial lag of standardized variable
  lag_var <- lag.listw(queen_weights, z_var)
  
  # Create classifications
  clusters <- rep("Not Significant", length(var))
  clusters[lisa > 0 & p_value <= p_threshold & z_var > 0 & lag_var > 0] <- "High-High"
  clusters[lisa > 0 & p_value <= p_threshold & z_var < 0 & lag_var < 0] <- "Low-Low"
  clusters[lisa < 0 & p_value <= p_threshold & z_var > 0 & lag_var < 0] <- "High-Low"
  clusters[lisa < 0 & p_value <= p_threshold & z_var < 0 & lag_var > 0] <- "Low-High"
  
  return(factor(clusters, levels = c("High-High", "Low-Low", "High-Low", "Low-High", "Not Significant")))
}

# Create LISA clusters for crime rates
columbus_sf$lisa_cluster <- lisa_clusters(columbus_sf$CRIME, 
                                         columbus_sf$local_moran, 
                                         columbus_sf$local_moran_p)

# Map the LISA clusters
tm_shape(columbus_sf) +
  tm_fill("lisa_cluster", 
          palette = c("red", "blue", "pink", "lightblue", "white"),
          title = "LISA Cluster") +
  tm_borders() +
  tm_layout(title = "LISA Cluster Map of Crime Rates",
            legend.outside = TRUE)
```

The LISA cluster map identifies:
- **High-High clusters (red)**: High-crime areas surrounded by high-crime areas (hotspots)
- **Low-Low clusters (blue)**: Low-crime areas surrounded by low-crime areas (coldspots)
- **High-Low outliers (pink)**: High-crime areas surrounded by low-crime areas
- **Low-High outliers (light blue)**: Low-crime areas surrounded by high-crime areas
- **Not significant (white)**: Areas where the pattern is not statistically significant

This integrated approach can help policymakers:

1. Identify areas with significant clustering of high crime rates
2. Prioritize interventions in areas that show consistent patterns across different methods
3. Target resources more effectively based on spatial patterns
4. Design context-specific interventions based on local conditions

# 10. Advanced Topics and Considerations

## Modifiable Areal Unit Problem (MAUP)

The MAUP refers to how results of spatial analyses can be affected by the specific boundaries used to aggregate data. Different boundary definitions can lead to different patterns and conclusions.

## Edge Effects

Edge effects occur because areas at the boundary of the study region have fewer neighbors, which can affect the calculation of spatial statistics.

## Statistical Significance vs. Practical Significance

Statistical significance doesn't always translate to practical importance. Consider the magnitude of the effects alongside p-values.

## Spatio-temporal Autocorrelation

Real-world processes often have both spatial and temporal dimensions. Methods exist to analyze spatio-temporal autocorrelation.

# 11. Conclusion

In this tutorial, we've explored:

1. **Global spatial autocorrelation** using Moran's I to detect overall clustering patterns
2. **Local spatial autocorrelation** using LISA and Getis-Ord Gi* to identify specific clusters and hotspots
3. **Different spatial weight structures** and their effects on the analysis
4. **Bivariate spatial relationships** between different variables
5. **Practical application** of these methods for policy interventions

Spatial autocorrelation analysis provides a powerful framework for understanding spatial patterns, identifying clusters, and informing targeted interventions. By applying these techniques, researchers and policymakers can gain deeper insights into spatial processes and make more informed decisions.
