---
title: "Geographically weighted regression models:"
subtitle: "A tutorial using the spgwr package in R"
author: "Carlos Mendez"
---
Suggested citation: 

> Mendez C. (2020).  Geographically weighted regression models: A tutorial using the spgwr package in R. R Studio/RPubs. Available at <https://rpubs.com/quarcs-lab/tutorial-gwr1>

This work is licensed under the Creative Commons Attribution-Share Alike 4.0 International License. 

Acknowledgment:

Material adapted from multiple sources, in particular the course materials of [Guy Lansley & James Cheshire (2016)](https://data.cdrc.ac.uk/tutorial/an-introduction-to-spatial-data-analysis-and-visualisation-in-r).

# Libraries

```{r message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)  # Modern data science workflow
library(sf)
library(sp)
library(tmap)
library(tmaptools)
library(spgwr)
library(grid)
library(gridExtra)

# Change the presentation of decimal numbers to 4 and avoid scientific notation
options(prompt="R> ", digits=4, scipen=999)
```

# Preliminary materials

- [Introduction to locally weighted regressions](https://youtu.be/Vf7oJ6z2LCc)
- [Introduction to Geographically Weighted Regressions](https://youtu.be/CpH8B2SiqdM) 

# Import datasets

 All used datasets can be obtained from [Guy Lansley & James Cheshire (2016)](https://data.cdrc.ac.uk/tutorial/an-introduction-to-spatial-data-analysis-and-visualisation-in-r).

## Non-spatial data

```{r}
Census.Data <-read.csv("practicaldata.csv")
```

```{r}
Census.Data[1:3,]
```


## Spatial data

```{r}
Output.Areas <- st_read(".", "Camden_oa11")
```


# Merge datasets
```{r}
OA.Census <- merge(Output.Areas, Census.Data, by.x="OA11CD", by.y="OA")
```

# Run linear model
Let us  run a linear model to understand the global relationship between our variables in our study area.

- Dependent variable:  The percentage of people with qualifications
- Independent (predictor) variables: the percentages of unemployed economically active adults and White British ethnicity 

```{r}
model <- lm(Qualification ~ Unemployed + White_British, data = Census.Data)
summary(model)

```

## Model diagnostics

This model has an adjusted R-squared value of 0.463. So we can assume 46% of the variance can be explained by the model. We can also observe the influences of each of the variables. 

A residual is the difference between the predicted and observed values for an observation in the model. **Models with lower r-squared values would have greater residuals on average** as the data would not fit the modeled regression line as well. Standardized residuals are represented as Z-scores where 0 represent the predicted values.

# Map residuals

Map the spatial distribution of the residuals

- Create a column vector of the residuals
```{r}
resids <- residuals(model)
summary(resids)
```

- Add the residuals column to the extended-spatial dataset

```{r}
map.resids <- cbind(OA.Census, resids) 
```

-  map the residuals using the quickmap function from tmap

```{r}
qtm(map.resids, fill = "resids")
```
If there is a geographic pattern in the residuals, it is possible that an unobserved variable may be influencing the dependent variable

# Geographically weighted regression

## Calculate kernel bandwidth

- Use an adaptive kernel

```{r}
ar_OA.Census <- as(OA.Census, "Spatial")
GWRbandwidth <- gwr.sel(Qualification ~ Unemployed + White_British, data = ar_OA.Census, adapt = T)
```
## Run the GWR model

```{r}
gwr.model = gwr(Qualification ~ Unemployed + White_British,
                data = ar_OA.Census,
                adapt=GWRbandwidth,
                hatmatrix=TRUE,
                se.fit=TRUE) 
```

Print results

```{r}
gwr.model
```

## Create results dataframe 

```{r}
results <-as.data.frame(gwr.model$SDF)
names(results)
```

## Map results

- Bind the results to `OA.Census` polygon.

```{r}
gwr.map <- cbind(OA.Census, as.matrix(results))
```

```{r}
gwr.map2 <- st_as_sf(gwr.map)
```


The variable names followed by the name of our original data frame (i.e. OA.Census.Unemployed) are the local R-Square of the model.

```{r}
qtm(gwr.map, fill = "localR2")
```

- Spatial distribution of White_British

```{r}
map1 <- tm_shape(gwr.map2) + 
  tm_fill("White_British",
          n = 5,
          style = "quantile")  +
  tm_layout(frame = FALSE,
            legend.text.size = 0.5,
            legend.title.size = 0.6)
map1
```

- Coefficients of  White_British


```{r}
map2 <- tm_shape(gwr.map2) +
  tm_fill("White_British.1",
          n = 5,
          style = "quantile",
          title = "WB Coefficient") +
  tm_layout(frame = FALSE,
            legend.text.size = 0.5,
            legend.title.size = 0.6)
map2
```


- Spatial distribution of Unemployed 

```{r}
map3 <- tm_shape(gwr.map) +
  tm_fill("Unemployed",
          n = 5,
          style = "quantile") +
  tm_layout(frame = FALSE,
            legend.text.size = 0.5,
            legend.title.size = 0.6)
map3
```

- Coefficients of Unemployed

```{r}
map4 <- tm_shape(gwr.map) +
  tm_fill("Unemployed.1",
          n = 5,
          style = "quantile",
          title = "Ue Coefficient") +
  tm_layout(frame = FALSE,
            legend.text.size = 0.5,
            legend.title.size = 0.6)
map4 
```



# Use Grid Extra

```{r}
# creates a clear grid
grid.newpage()

# assigns the cell size of the grid, in this case 2 by 2
pushViewport(viewport(layout=grid.layout(2,2)))

# prints a map object into a defined cell   
print(map1, vp=viewport(layout.pos.col = 1, layout.pos.row =1))
print(map2, vp=viewport(layout.pos.col = 2, layout.pos.row =1))
print(map3, vp=viewport(layout.pos.col = 1, layout.pos.row =2))
print(map4, vp=viewport(layout.pos.col = 2, layout.pos.row =2))
```
